---
title: "Evaluate the cell type deconvolution performance of Seurat integration"
author: "Ningshan Li"
date: "2024/09/11"
output: 
  html_notebook:
    code_folding: hide
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, results='hold', fig.width = 7, fig.height = 5, dpi = 300)


# use ggpubr package (http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/81-ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page/)
library(dplyr)
library(ggplot2)


`%notin%` = Negate(`%in%`)

set.seed(1)


home.dir = '/home/hill103/Documents/Spatial/Figures/Simulation'
save_file = FALSE

my_color = c('#e6194b', '#3cb44b', '#ffe119', '#4363d8', '#f58231', '#911eb4', '#46f0f0', '#f032e6', '#bcf60c', '#fabebe', '#008080', '#e6beff', '#9a6324', '#fffac8', '#800000', '#aaffc3', '#808000', '#ffd8b1', '#000075', '#808080', '#ffffff', '#000000')
method_color = c("SDePER"='#E6194B', "GLRM"='#F032E6', "RCTD"='#46F0F0', "SpatialDWLS"='#3CB44B', "cell2location"='#FFE119',"SONAR"='#9A6324', "SPOTlight"='#4363D8', "CARD"='#F58231', "DestVI"='#911EB4', "Seurat+GLRM"='#808080')
method_order = names(method_color)
```


# Summary

This R Notebook generates boxplots of performance to evaluate the cell type deconvolution performance of **Seurat integration** (version 4.3) + GLRM modeling.

1. **Inputs**:

  * [`simulation_all_results.rds`](https://github.com/az7jh2/SDePER_Analysis/blob/main/Figures/Simulation/simulation_all_results.rds): cell type deconvolution results of all methods, as well as the ground truth in simulation analysis.

2. **Outputs**:

  * Table. Median RMSE, JSD, correlation and FDR in Scenario 1
  * Figure. Boxplot of performance in Scenario 1


# Version

```{r}
version[['version.string']]
print(sprintf('Package %s version: %s', 'ggplot2', packageVersion('ggplot2')))
print(sprintf('Package %s version: %s', 'ggpubr', packageVersion('ggpubr')))
print(sprintf('Package %s version: %s', 'philentropy', packageVersion('philentropy'))) # JSD function
```


# Read estimated cell type proportions by all methods

Here we only evaluate results of SDePER, GLRM and Seurat+GLRM.

```{r}
file_name = file.path(home.dir, 'simulation_all_results.rds')
all_res = readRDS(file_name)
print(sprintf('load data from %s', file_name))

# check the order of spatial spots and cell types are consistent
all_spots = row.names(all_res[['Truth']])
all_celltypes_S1 = colnames(all_res[['Truth']])
all_methods = c('SDePER', 'GLRM', 'Seurat+GLRM')
```


Check the order of spots and cell types are consistent before performance evaluation.

```{r}
for (method_name in all_methods) {
  stopifnot(all(row.names(all_res[['S1_ext_ref']][[method_name]]) == all_spots))
  stopifnot(all(colnames(all_res[['S1_ext_ref']][[method_name]]) == all_celltypes_S1))
}
```


Check whether negative values of estimated cell type proportions exist, as negative values may cause error in JSD calculation and got `NaN`. Replace them as 0.

```{r}
for (method_name in all_methods) {
  tmp_df = all_res[['S1_ext_ref']][[method_name]]
  for (i in 1:nrow(tmp_df)) {
    for (j in 1:ncol(tmp_df)) {
      if (tmp_df[i, j] < 0) {
        print(sprintf('%s: %s result: row %d (%s) column %d (%s) has negative value %g', 'S1_ext_ref', method_name, i, row.names(tmp_df)[i], j, colnames(tmp_df)[j], tmp_df[i, j]))
        # replace them with 0
        all_res[['S1_ext_ref']][[method_name]][i, j] = 0
      }
    }
  }
}
```

Check whether estimated cell type proportions of spot are ALL 0s.

```{r}
for (method_name in all_methods) {
  tmp_df = all_res[['S1_ext_ref']][[method_name]]
  for (i in 1:nrow(tmp_df)) {
    if (sum(tmp_df[i, ]) == 0) {
      print(sprintf('%s: %s result: row %d (%s) has ALL 0s', 'S1_ext_ref', method_name, i, row.names(tmp_df)[i]))
    }
  }
}
```



# Evaluate performance of cell type deconvolution methods

## Calculate spot-wise performance of all methods

4 performance measurements:

* root mean square error (**RMSE**): quantifies the overall estimation accuracy
* Jensen-Shannon Divergence (**JSD**): assesses similarity between the estimated cell type distribution and ground-truth per spot
* **Pearson’s correlation coefficient**: measures the similarity of estimation to ground-truth
* false discovery rate (**FDR**): measures how many cell types were falsely predicted to be present


```{r}
binaryPredEvaluation = function(truth, pred) {
  # Given an array of truth and predictions (either 0 or 1), calculate confusion matrix
  # convert to factors to ensure get a full 2*2 confusion matrix
  truth_factor = factor(truth, levels = c(0, 1))
  pred_factor = factor(pred, levels = c(0, 1))
  # Generate confusion matrix
  conf_matrix = table(Actual = truth_factor, Predicted = pred_factor)
  # Extract elements of the confusion matrix
  TP = conf_matrix[2, 2]
  FP = conf_matrix[1, 2]
  FN = conf_matrix[2, 1]
  TN = conf_matrix[1, 1]
  # Calculate FDR (False Discovery Rate)
  FDR = FP / (TP + FP)
  # Calculate FNR (False Negative Rate)
  FNR = FN / (TP + FN)
  return(c(FDR, FNR))
}

calcPerformance = function(truth, pred) {
  # calculate RMSE, JSD, correlation and FDR for each row
  # inputs are matrix with rows as spots and columns as cell types, order has been checked to be consistent
  stopifnot(all(row.names(truth) == row.names(pred)))
  stopifnot(all(colnames(truth) == colnames(pred)))
  
  # binary cell type proportions (0:absent; 1:present)
  truth_binary = truth != 0
  pred_binary = pred != 0
  # in-place conversion from bool to 0/1 while keeping dimensions, row names and column names, as `as.numeric()` will "flattern" the original matrix
  truth_binary[] = as.numeric(truth_binary)
  pred_binary[] = as.numeric(pred_binary)
  
  perform_df = data.frame(matrix(ncol=5, nrow=0))
  colnames(perform_df) = c('RMSE', 'JSD', 'Pearson', 'FDR', 'FNR')
  
  for (i in 1:nrow(truth)) {
    RMSE = sqrt(mean((truth[i,] - pred[i,]) ^ 2))
    if (sum(pred[i,])>0 & sum(truth[i,])>0) {
      JSD = philentropy::JSD(rbind(truth[i,], pred[i,]), unit = 'log2', est.prob = 'empirical')
    } else {
      JSD = 1
    }
    Pearson = cor.test(truth[i,], pred[i,])$estimate
    tmp = binaryPredEvaluation(truth_binary[i,], pred_binary[i,])
    FDR = tmp[1]
    FNR = tmp[2]
    
    perform_df[nrow(perform_df)+1, ] = c(RMSE, JSD, Pearson, FDR, FNR)
  }
  
  # also record spot names
  stopifnot(nrow(perform_df) == nrow(truth))
  perform_df['Spot'] = row.names(truth)
  
  return(perform_df)
}


all_perform = list()

all_perform[['S1_ext_ref']] = list()

for (method_name in all_methods) {
  all_perform[['S1_ext_ref']][[method_name]] = calcPerformance(as.matrix(all_res[['Truth']]), as.matrix(all_res[['S1_ext_ref']][[method_name]]))
}
```


## Summary spot-wise performance into method-wise

Including Table. Median RMSE, JSD, correlation and FDR in Scenario 1.

```{r}
perform_raw_df = data.frame(matrix(ncol=9, nrow=0))
colnames(perform_raw_df) = c('Scenario', 'Method', 'Reference', 'Spot', 'RMSE', 'JSD', 'Pearson', 'FDR', 'FNR')

# calculate median performance across all spatial spots for all methods
perform_median_df = data.frame(matrix(ncol=8, nrow=0))
colnames(perform_median_df) = c('Scenario', 'Method', 'Reference', 'median_RMSE', 'median_JSD', 'median_Pearson', 'median_FDR', 'median_FNR')

this_scenario = 'Scenario 1'
this_ref = 'External'
 
for (method_name in all_methods) {
  tmp_df = all_perform[['S1_ext_ref']][[method_name]]
  tmp_df['Scenario'] = this_scenario
  tmp_df['Method'] = method_name
  tmp_df['Reference'] = this_ref
  
  perform_raw_df = rbind(perform_raw_df, tmp_df[, c('Scenario', 'Method', 'Reference', 'RMSE', 'JSD', 'Pearson', 'FDR', 'FNR')])
  
  perform_median_df[nrow(perform_median_df)+1, ] = c(this_scenario, method_name, this_ref,
                                                     round(median(tmp_df$RMSE), 3),
                                                     round(median(tmp_df$JSD), 3),
                                                     round(median(tmp_df$Pearson), 3),
                                                     round(median(tmp_df$FDR), 3),
                                                     round(median(tmp_df$FNR), 3))
}

# set method column as factors
perform_raw_df['Method'] = factor(perform_raw_df$Method, levels = all_methods)

perform_median_df[, c('Scenario', 'Method', 'Reference', 'median_RMSE', 'median_JSD', 'median_Pearson', 'median_FDR', 'median_FNR')]
```



# Draw figures

## Figure. Boxplot of performance in Scenario 1

```{r, fig.width=4, fig.height=6}
if (save_file) {
  file_name = file.path(home.dir, 'FigX_S1_performance.pdf')
  cairo_pdf(file_name, height=6, width=4, onefile=T)
  print(sprintf('figures saved in file %s', file_name))
}

plot_df = perform_raw_df %>%
  filter(Scenario=='Scenario 1')

g_list = list()

for (perform_ind in c('RMSE', 'Pearson', 'JSD', 'FDR')) {
  g_list[[perform_ind]] = ggplot(plot_df, aes(x=Method, y=.data[[perform_ind]], fill=Method)) +
                            geom_boxplot(position=position_dodge(), outlier.shape=NA) +
                            scale_fill_manual(values=method_color) +
                            theme_classic() +
                            theme(strip.text = element_text(size=10),
                                  axis.text = element_text(color="black"),
                                  axis.ticks.x = element_blank(),
                                  axis.text.x = element_blank(),
                                  axis.title.x = element_blank(),
                                  legend.title = element_blank())
}

g_list[['Pearson']] = g_list[['Pearson']] + geom_hline(yintercept=0, color="red", linetype="dashed")

ggpubr::ggarrange(plotlist=g_list, ncol=2, nrow=2, common.legend=TRUE, legend="right")
```


## Addtional boxplot for FNR

```{r, fig.width=3, fig.height=3}
ggplot(plot_df, aes(x=Method, y=FNR, fill=Method)) +
  geom_boxplot(position=position_dodge(), outlier.shape=NA) +
  scale_fill_manual(values=method_color) +
  theme_classic() +
  theme(strip.text = element_text(size=10),
        axis.text = element_text(color="black"),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        legend.title = element_blank())
```


