---
title: "Generate figures in *Sequencing-based* Simulation Analysis"
author: "Ningshan Li & Yunqing Liu"
date: "2024/05/09"
output: 
  html_notebook:
    code_folding: hide
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, results='hold', fig.width = 7, fig.height = 5, dpi = 300)


# use ggpubr package (http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/81-ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page/)
library(dplyr)
library(ggplot2)
library(scatterpie)


`%notin%` = Negate(`%in%`)

set.seed(1)

home.dir = '/home/hill103/Documents/Spatial/Figures/Simulation_seq_based'
save_file = FALSE

my_color = c('#e6194b', '#3cb44b', '#ffe119', '#4363d8', '#f58231', '#911eb4', '#46f0f0', '#f032e6', '#bcf60c', '#fabebe', '#008080', '#e6beff', '#9a6324', '#fffac8', '#800000', '#aaffc3', '#808000', '#ffd8b1', '#000075', '#808080', '#ffffff', '#000000')
method_color = c("SDePER"='#E6194B', "GLRM"='#F032E6', "RCTD"='#46F0F0', "SpatialDWLS"='#3CB44B', "cell2location"='#FFE119',"SONAR"='#9A6324', "SPOTlight"='#4363D8', "CARD"='#F58231', "DestVI"='#911EB4')
```


# Summary

This R Notebook generates figures in **Sequencing-based Simulation Analysis** section in manuscript.

1. **Inputs**:

    * [`simulation_seq_based_all_results.rds`](https://github.com/az7jh2/SDePER_Analysis/blob/main/Figures/Simulation_seq_based/simulation_seq_based_all_results.rds): cell type deconvolution results of all methods, as well as the ground truth in sequencing-based simulation analysis.
    * [`sim_spatial_spot_loc.csv`](https://github.com/az7jh2/SDePER_Analysis/blob/main/Simulation/Generate_simulation_data/sim_spatial_spot_loc.csv): physical locations of spatial spots.

2. **Outputs**:

    * Table. Median RMSE, JSD, correlation and FDR of all methods in Scenario 1
    * Figure. Boxplot of performance of all methods in Scenario 1
    * Figure. Pie chart of cell type proportions of all methods in Scenario 1


# Version

```{r}
version[['version.string']]
print(sprintf('Package %s version: %s', 'ggplot2', packageVersion('ggplot2')))
print(sprintf('Package %s version: %s', 'ggpubr', packageVersion('ggpubr')))
print(sprintf('Package %s version: %s', 'philentropy', packageVersion('philentropy'))) # JSD function
```

# Read locations of spatial spots

```{r}
file_name = file.path(home.dir, 'sim_spatial_spot_loc.csv')
loc_df = read.csv(file_name, row.names = 1, check.names = F)
print(sprintf('load data from %s', file_name))
```


# Read estimated cell type proportions by all methods

```{r}
file_name = file.path(home.dir, 'simulation_seq_based_all_results.rds')
all_res = readRDS(file_name)
print(sprintf('load data from %s', file_name))
```


# Quality control

## Variable order

Check the order of spots and cell types are consistent before performance evaluation.

```{r}
for (scenario in names(all_res)) {
  if (scenario != 'Truth') {
    for (method_name in names(all_res[[scenario]])) {
      stopifnot(all(row.names(all_res[[scenario]][[method_name]]) == row.names(all_res$Truth)))
      stopifnot(all(colnames(all_res[[scenario]][[method_name]]) == colnames(all_res$Truth)))
    }
  }
}
```


## Negative proportions

Check whether negative values of estimated cell type proportions exist, as negative values may cause error in JSD calculation and got `NaN`. Replace them as 0.

```{r}
for (scenario in names(all_res)) {
  if (scenario != 'Truth') {
    for (method_name in names(all_res[[scenario]])) {
      tmp_df = all_res[[scenario]][[method_name]]
      for (i in 1:nrow(tmp_df)) {
        for (j in 1:ncol(tmp_df)) {
          if (tmp_df[i, j] < 0) {
            print(sprintf('%s: %s result: row %d (%s) column %d (%s) has negative value %g', scenario, method_name, i, row.names(tmp_df)[i], j, colnames(tmp_df)[j], tmp_df[i, j]))
            # replace them with 0
            all_res[[scenario]][[method_name]][i, j] = 0
          }
        }
      }
    }
  }
}
```

## ALL 0s

Check whether estimated cell type proportions of spot are ALL 0s.

```{r}
for (scenario in names(all_res)) {
  if (scenario != 'Truth') {
    for (method_name in names(all_res[[scenario]])) {
      tmp_df = all_res[[scenario]][[method_name]]
      for (i in 1:nrow(tmp_df)) {
        if (sum(tmp_df[i, ]) == 0) {
          print(sprintf('%s: %s result: row %d (%s) has ALL 0s', scenario, method_name, i, row.names(tmp_df)[i]))
        }
      }
    }
  }
}
```

## Hard-thresholding (NOT applied)

For the methods without cell type selection procedure, including **cell2location**, **DestVI**, **CARD** and **SONAR**, we put a hard-thresholding with cutoff **0.1** on the estimated cell type compositions to force the extremely small proportions to be zeros.

**UPDATE**: hard-thresholding were NOT applied, original estimated cell type proportions were used for performance evaluation.



# Evaluate performance of cell type deconvolution methods

## Calculate spot-wise performance of all methods

4 performance measurements:

* root mean square error (**RMSE**): quantifies the overall estimation accuracy
* Jensen-Shannon Divergence (**JSD**): assesses similarity between the estimated cell type distribution and ground-truth per spot
* **Pearson’s correlation coefficient**: measures the similarity of estimation to ground-truth
* false discovery rate (**FDR**): measures how many cell types were falsely predicted to be present

NOTE:

* in **SpatialDWLS** result, rows with ALL 0s may cause performance values to be `NA` or `NaN`.

```{r}
binaryPredEvaluation = function(truth, pred) {
  # Given an array of truth and predictions (either 0 or 1), calculate confusion matrix
  # convert to factors to ensure get a full 2*2 confusion matrix
  truth_factor = factor(truth, levels = c(0, 1))
  pred_factor = factor(pred, levels = c(0, 1))
  # Generate confusion matrix
  conf_matrix = table(Actual = truth_factor, Predicted = pred_factor)
  # Extract elements of the confusion matrix
  TP = conf_matrix[2, 2]
  FP = conf_matrix[1, 2]
  FN = conf_matrix[2, 1]
  TN = conf_matrix[1, 1]
  # Calculate FDR (False Discovery Rate)
  FDR = FP / (TP + FP)
  # Calculate FNR (False Negative Rate)
  FNR = FN / (TP + FN)
  return(c(FDR, FNR))
}

calcPerformance = function(truth, pred) {
  # calculate RMSE, JSD, correlation and FDR for each row
  # inputs are matrix with rows as spots and columns as cell types, order has been checked to be consistent
  stopifnot(all(row.names(truth) == row.names(pred)))
  stopifnot(all(colnames(truth) == colnames(pred)))
  
  # binary cell type proportions (0:absent; 1:present)
  truth_binary = truth != 0
  pred_binary = pred != 0
  # in-place conversion from bool to 0/1 while keeping dimensions, row names and column names, as `as.numeric()` will "flattern" the original matrix
  truth_binary[] = as.numeric(truth_binary)
  pred_binary[] = as.numeric(pred_binary)
  
  perform_df = data.frame(RMSE=numeric(), JSD=numeric(), Pearson=numeric(), FDR=numeric(), FNR=numeric(), stringsAsFactors = F)
  
  for (i in 1:nrow(truth)) {
    RMSE = sqrt(mean((truth[i,] - pred[i,]) ^ 2))
    if (sum(pred[i,])>0 & sum(truth[i,])>0) {
      JSD = philentropy::JSD(rbind(truth[i,], pred[i,]), unit = 'log2', est.prob = 'empirical')
    } else {
      JSD = 1
    }
    Pearson = cor.test(truth[i,], pred[i,])$estimate
    tmp = binaryPredEvaluation(truth_binary[i,], pred_binary[i,])
    FDR = tmp[1]
    FNR = tmp[2]
    
    perform_df[nrow(perform_df)+1, ] = c(RMSE, JSD, Pearson, FDR, FNR)
  }
  
  # also record spot names
  stopifnot(nrow(perform_df) == nrow(truth))
  perform_df['Spot'] = row.names(truth)
  
  return(perform_df)
}


all_perform = list()

for (scenario in names(all_res)) {
  if (scenario != 'Truth') {
    all_perform[[scenario]] = list()
    for (method_name in names(all_res[[scenario]])) {
      all_perform[[scenario]][[method_name]] = calcPerformance(as.matrix(all_res[['Truth']]), as.matrix(all_res[[scenario]][[method_name]]))
    }
  }   
}
```


For rows with ALL 0s in **SpatialDWLS** performance, replace `NA` Pearson’s correlation coefficient as 0, `NaN` FDR as 0.

```{r}
for (scenario in names(all_perform)) {
  tmp_df = all_perform[[scenario]][['SpatialDWLS']]
  tmp_df[is.na(tmp_df)] = 0
  all_perform[[scenario]][['SpatialDWLS']] = tmp_df
}
```



## Summary spot-wise performance into method-wise

Including Table. Median RMSE, JSD, correlation, FDR and FNR of all methods in Scenario 1.

```{r}
perform_raw_df = data.frame(Scenario=character(), Method=character(), Reference=character(), Spot=character(),
                            RMSE=numeric(), JSD=numeric(), Pearson=numeric(), FDR=numeric(), FNR=numeric(),
                            stringsAsFactors = F)

# calculate median performance across all spatial spots for all methods
perform_median_df = data.frame(Scenario=character(), Method=character(), Reference=character(),
                               median_RMSE=numeric(), median_JSD=numeric(), median_Pearson=numeric(), median_FDR=numeric(), median_FNR=numeric(),
                               stringsAsFactors = F)

for (scenario in names(all_perform)) {
  if (scenario == "S1_int_ref") {
    this_scenario = 'Scenario 1'
    this_ref = 'Internal'
  } else if (scenario == "S1_ext_ref") {
    this_scenario = 'Scenario 1'
    this_ref = 'External'
  } else if (scenario == 'S2_ext_ref') {
    this_scenario = 'Scenario 2'
    this_ref = 'External'
  } else {
    this_scenario = 'Scenario 3'
    this_ref = 'External'
  }
  
  for (method_name in names(all_perform[[scenario]])) {
    tmp_df = all_perform[[scenario]][[method_name]]
    tmp_df['Scenario'] = this_scenario
    tmp_df['Method'] = method_name
    tmp_df['Reference'] = this_ref
    
    perform_raw_df = rbind(perform_raw_df, tmp_df[, c('Scenario', 'Method', 'Reference', 'Spot', 'RMSE', 'JSD', 'Pearson', 'FDR', 'FNR')])
    
    # use list() instead of c() to keep data type, otherwise c() will coerce all elements to a common type like string
    perform_median_df[nrow(perform_median_df)+1, ] = list(this_scenario, method_name, this_ref,
                                                          round(median(tmp_df$RMSE), 3),
                                                          round(median(tmp_df$JSD), 3),
                                                          round(median(tmp_df$Pearson), 3),
                                                          round(median(tmp_df$FDR), 3),
                                                          round(median(tmp_df$FNR), 3))
  }
}

# set method column as factors
perform_raw_df['Method'] = factor(perform_raw_df$Method, levels = names(method_color))

perform_median_df[, c('Scenario', 'Method', 'Reference', 'median_RMSE', 'median_JSD', 'median_Pearson', 'median_FDR', 'median_FNR')]
```



# Draw figures

## Figure. Boxplot of performance of all methods in Scenario 1

```{r, fig.width=6.5, fig.height=6}
if (save_file) {
  file_name = file.path(home.dir, 'Fig_sequencing_based_performance.pdf')
  cairo_pdf(file_name, height=6, width=6.5, onefile=T)
  print(sprintf('figures saved in file %s', file_name))
}

plot_df = perform_raw_df %>%
  filter(Scenario=='Scenario 1')

g_list = list()

for (perform_ind in c('RMSE', 'Pearson', 'JSD', 'FDR')) {
  g_list[[perform_ind]] = ggplot(plot_df, aes(x=Method, y=.data[[perform_ind]], fill=Method)) +
                            geom_boxplot(position=position_dodge(), outlier.shape=NA) +
                            scale_fill_manual(values=method_color) +
                            theme_classic() +
                            theme(strip.text = element_text(size=10),
                                  axis.text = element_text(color="black"),
                                  axis.ticks.x = element_blank(),
                                  axis.text.x = element_blank(),
                                  axis.title.x = element_blank(),
                                  legend.title = element_blank()) +
                            facet_grid(~Reference)
}

g_list[['Pearson']] = g_list[['Pearson']] + geom_hline(yintercept=0, color="red", linetype="dashed")

ggpubr::ggarrange(plotlist=g_list, ncol=2, nrow=2, common.legend=TRUE, legend="right")
```


## Addtional boxplot for FNR

```{r, fig.width=4, fig.height=3}
ggplot(plot_df, aes(x=Method, y=FNR, fill=Method)) +
  geom_boxplot(position=position_dodge(), outlier.shape=NA) +
  scale_fill_manual(values=method_color) +
  theme_classic() +
  theme(strip.text = element_text(size=10),
        axis.text = element_text(color="black"),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        legend.title = element_blank()) +
  facet_grid(~Reference)
```


## Figure. Pie chart of all methods in Scenario 1

### Internal reference

```{r, fig.width=19.5, fig.height=21}
if (save_file) {
  file_name = file.path(home.dir, 'Fig_sequencing_based_piechart_internal_ref.pdf')
  cairo_pdf(file_name, height=21, width=19.5, onefile=T)
  print(sprintf('figures saved in file %s', file_name))
}

g_list = list()


all_methods = names(method_color)
all_cts = colnames(all_res$Truth)

for (one_method in all_methods) {
  plot_df = merge(all_res$S1_int_ref[[one_method]], loc_df[, c('x', 'y')], by='row.names')
  g_list[[length(g_list)+1]] = ggplot() +
    geom_scatterpie(aes(x=x, y=y), data=plot_df, cols=all_cts) +
    scale_fill_manual(values=my_color) +
    theme_classic() +
    ggtitle(one_method) +
    theme(axis.title=element_blank(), axis.text=element_blank(), axis.ticks=element_blank(), axis.line=element_blank(),
      legend.title=element_blank(), legend.text=element_text(size=15),
      plot.title=element_text(size=17, face='bold', hjust=0.1))
}

ggpubr::ggarrange(plotlist = g_list, ncol = 3, nrow = 3, align = 'hv', common.legend = T, legend = 'right', font.label = list(size=17))
```


### External reference

```{r, fig.width=19.5, fig.height=21}
if (save_file) {
  file_name = file.path(home.dir, 'Fig_sequencing_based_piechart_external_ref.pdf')
  cairo_pdf(file_name, height=21, width=19.5, onefile=T)
  print(sprintf('figures saved in file %s', file_name))
}

g_list = list()


all_methods = names(method_color)
all_cts = colnames(all_res$Truth)

for (one_method in all_methods) {
  plot_df = merge(all_res$S1_ext_ref[[one_method]], loc_df[, c('x', 'y')], by='row.names')
  g_list[[length(g_list)+1]] = ggplot() +
    geom_scatterpie(aes(x=x, y=y), data=plot_df, cols=all_cts) +
    scale_fill_manual(values=my_color) +
    theme_classic() +
    ggtitle(one_method) +
    theme(axis.title=element_blank(), axis.text=element_blank(), axis.ticks=element_blank(), axis.line=element_blank(),
      legend.title=element_blank(), legend.text=element_text(size=15),
      plot.title=element_text(size=17, face='bold', hjust=0.1))
}

ggpubr::ggarrange(plotlist = g_list, ncol = 3, nrow = 3, align = 'hv', common.legend = T, legend = 'right', font.label = list(size=17))
```


